### 2014-10-28 contract-language.tex

The syntax of our core contract language is shown in Fig.
~\ref{fig:contract-lang}. The language is based on first-order logic
(FOL), and admits prenex quantification over typed effect variables.
We use a special effect variable ($\cureff$) to denote the effect of
\emph{current operation} - the operation for which a contract is being
written. The type of an effect is simply the label of the operation
(eg: \cf{withdraw}) that engendered the effect. We admit disjuntion in
types to let an effect variable range over effects of multiple
operations. Since each replicated data type comes with a unique set of
operations (therefore, unique set of types for effects), our contract
language is implictly parameterized over a replicated data type
specification.
\GK{I think sec 2. is the right place to introduce the idea of rdt
spec via the bank account example.}

Quantifier-free propositions in our contract language are
conjunctions, disjunctions and implications of predicates expressing
relations between pairs of effect variables. Syntactic class of
relations is seeded with primitive $\visZ$, $\soZ$, and $\sameobjZ$
relations, and also admits derived relations that are expressible as
union, intersection, or transitive closure\footnote{Strictly speaking,
$R^{+}$ is not the transitive closure of $R$, as transitive closure is
not expressible in FOL. Instead, $R^{+}$ in our language denotes
\emph{a} superset of transitive closure of $R$. Formally, $R^{+}$ is
any relation $R'$ such that forall $x$, $y$, and $z$, a) $R(x,y)
\Rightarrow R'(x,y)$, and b) $R'(x,y) \conj R'(y,z) \Rightarrow
R'(x,z)$} of primitive relations. Commonly used derived relations are
the \emph{same object session order} ($\sooZ$), and the the
\emph{happens before order} ($\hboZ$) defined as following:
\begin{smathpar}
\begin{array}{lcl}
{\sf soo} & = &  {\sf so} \cap {\sf sameobj}\\
{\sf hbo} & = & ({\sf soo} \cup {\sf vis})^{+}\\
\end{array}
\end{smathpar}

Note that our contract language (stripped of its syntactic sugar) is a
carefully chosen subset of first-order logic that is known to be
decidable~\cite{epr}. In particular, for any two contracts ($\cv_1$
and $\cv_2$) in our language, the implication check $\cv_1 \Rightarrow
\cv_2$ is decidable, allowing us to automatically deduce that $\cv_2$
is weaker than $\cv_1$. We formally define the \emph{weaker than}
relation among contracts as following: 
\begin{definition}
A contract $\cv_2$ is said to be weaker than $\cv_1$ (denoted $\cv_2
\le \cv_1$ ) if and only if:
\begin{center}
$\Delta \vdash \cv_1 \Rightarrow \cv_2$
\end{center}
\end{definition}
\noindent The context $\Delta$ in the above definition captures assumptions
about the nature of primitive relations, such as the assumption that
$\Rvis$ and $\Rso$ are irreflexive.

This ability to statically compare contracts is key to contract
classification and subsequent enforcement. This is described in the
following subsection.

@@

Application-level consistency requirements expressed by \name
contracts can have varying impact on the available of the data store
trying to meet the requirements. We identify three major classes of
contracts that differ on this parameter: 
% For instance, total-order requirement
% expressed by the \cf{withdraw} contract (Sec. ~\ref{sec:motivation})
% imposes global coordination among all replicas, and consequently makes
% the store unavailable for any concurrent session. In contrast, the
% trivial (\cf{true}) contract of the \cf{deposit} operation expresses
% no special consistency requirements; any replica can serve a deposit
% request immediately, and the store remains available to concurrent
% sessions. Based on how a contract affects the availability of the data
% store, we classify it into one of the following three categories:

\begin{itemize}
\item \textbf{Unavailable}: These are the contracts that are
satisfiable only under strong consistency. The need for global
coordination to satisfy such contracts makes the store unavailable for
concurrent sessions. An example is the total-order contract on
\cf{withdraw} (Sec. ~\ref{sec:motivation}).
\item \textbf{Sticky-available}: These are the contracts that require
see a \emph{causally consistent} snapshot of the store that
\emph{must} include some or all of the operations from the same
session. By applying all operations of a session to the same logical
replica (colloquially, \emph{sticking} a session to a replica), it is
possible for a store to satisfy such contracts, while remaining highly
available. The \cf{getBalance} contract that is sufficient to prevent
the missing update anomaly is a sticky available contract.
\item \textbf{Highly-available}: Contracts that only require to see
\emph{some} causally consistent snapshop of the store are said to
require \emph{causal visiblity} guarantee. This guarantee can be
achieved by forcing every replica to reveal only those effects whose
preceding effects are already included in the replica. Consequently,
such contracts can always be satisfied by the store, letting it remain
highly available for other requests.
\end{itemize}

It is important to note that, although finer distinctions can be made
between availability implications of contracts, the above
classification scheme is the one that is known to have practical
significance\cite{bailisHAT}. More importantly, the consistency levels
described in the classification scheme, namely strong consistency
({\sf sc}), causal consistency ({\sc cc}), and causal visibility ({\sf
cv}), can be implemented efficiently on most off-the-shelf eventually
consistent data stores\cite{...}. By determining which of the
aformentioned classes a contract falls into, the corresponding
implementation of the consistency level on the data store can be used
to satisfy the contract efficiently. Nonetheless, our classification
scheme is extendible, meaning that a new availability class can be
carved out of existing ones with the advent of new data stores
supporting efficient enforcement of finer consistency levels.

To classify contracts into one of the aforementioned availability
classes, we need to determine if the corresponding consistency
guarantee is sufficient to satisfy the contract. Towards this end, we
encode the {\sf sc}, {\sf cc}, and {\sf cv} consistency guarantees as
formulas in our contract language ($\hboZ$, the happens-before order,
was defined previously in the same section):
\begin{smathpar}
\stretcharraybig
\begin{array}{rcl}
{\sf sc} & = 		 & \forall a. (\hbo{a}{\cureff} \vee \hbo{\cureff}{a} \vee a = \cureff) \\
         & & \conj (\hbo{a}{\cureff} \Rightarrow \vis{a}{\cureff}) \\
				 & & \conj (\hbo{\cureff}{a} \Rightarrow \vis{\cureff}{a}) \\
{\sf cc} & = & \forall a. \hbo{a}{\cureff} \Rightarrow \vis{a}{\cureff} \\
{\sf cv} & = & \forall a,b. \hbo{a}{b} \wedge \vis{b}{\cureff} \Rightarrow \vis{a}{\cureff} \\
\end{array}
\end{smathpar}
Comparing ($\le$) a contract formula against the above formulas of
{\sf sc}, {\sf cc}, and {\sf cv} will let us determine if the contract
can be satisfied under strong consistency, causal consistency, and
causal visiblity consistency guarantees, respectively. Note that {\sf
sc}, {\sf cc} and {\sf cv} are themselves totally ordered with respect
to the $\le$ relation:
\begin{center}
$\cvc \le \ccc \le \scc$
\end{center}
This concurs with the intuition that any contract satisfiable under
$\ccc$ or $\cvc$ is satisfiable under $\scc$ , and any contract that
is satisfiable under $\cvc$ is satisfiable under $\ccc$. Nonetheless,
we determine the availability class of a contract based on the
\emph{weakest} guarantee (among $\cvc$, $\ccc$, and $\scc$) required
to satisfy the contract. The classification scheme is presented
formally as rules in Figure ~\ref{sem:classfiy}.
% Contract classification rules
% ------------------------------
\input{contract-classification}
The rules to classify contracts into availability classes are
straightforward. The remaining rule judges a contract as {\sf
WellFormed} if and only if it can be satisfied under strong
consistency. We only consider well-formed contracts for classification
and subsequent enforcement at run-time. The rest are reported as
errors.

As a result of our automatic classification scheme, a store that can
enforce $\scc$, $\ccc$, and $\cvc$ contracts can also enforce any
well-formed \name contract by determining its availability class.  The
following section formally describes operational semantics of one such
store, along with a meta-theoretic result that certifies the soundness
of our classification-based contract enforcement strategy.

>>>>

The syntax of our core contract language is shown in Fig.
~\ref{fig:contract-lang}. The language is based on first-order logic (FOL), and
admits prenex quantification over typed effect variables. We use a special
effect variable ($\cureff$) to denote the effect of \emph{current operation} -
the operation for which a contract is being written. The type of an effect is
simply the name of the operation (eg: \cf{withdraw}) that engendered the
effect. We admit disjuntion in types to let an effect variable range over
multiple operation names.

Quantifier-free propositions in our contract language are conjunctions,
disjunctions and implications of predicates expressing relations between pairs
of effect variables. Syntactic class of relations is seeded with primitive
$\visZ$, $\soZ$, and $\sameobjZ$ relations, and also admits derived relations
that are expressible as union, intersection, or transitive
closure\footnote{Strictly speaking, $R^{+}$ is not the transitive closure of
$R$, as transitive closure is not expressible in FOL. Instead, $R^{+}$ in our
language denotes \emph{a} superset of transitive closure of $R$. Formally,
$R^{+}$ is any relation $R'$ such that forall $x$, $y$, and $z$, a) $R(x,y)
\Rightarrow R'(x,y)$, and b) $R'(x,y) \conj R'(y,z) \Rightarrow R'(x,z)$} of
primitive relations. Commonly used derived relations are the \emph{same object
session order} ($\small \sooZ = \soZ ~\cap~ \sameobjZ$), the \emph{happens
before order} ($\small \hbZ = (\soZ ~\cup~ \visZ)^+$), and the per-object
happens-before order ($\small \hboZ = (\sooZ ~\cup~ \visZ)^+$).

\subsection{Capturing Store Semantics}
\label{sec:store_sem}

An important aspect of our contract classification system is that the store
semantics is also captured using the same contract language used to describe
application-level consistency. In this regard, similar
to~\cite{Burckhardt2014}, we can rigorously define a wide variety of store
semantics including those that combine any subset of session and causality
guarantees, and multiple consistency levels. For example, a store that offers
strong consistency is captured by the contract:

\vspace{-1em}
\begin{smathpar}
\scc = \forall a.~\sameobj{a}{\cureff} \Rightarrow \vis{a}{\cureff} ~\vee~ \vis{\cureff}{a} ~\vee~ a = \cureff
\end{smathpar}

Similarly, a store that offers per-object causal consistency is captured by the
contract:

\vspace{-1em}
\begin{smathpar}
\ccc = \forall a.~(\hboZ \cap \sameobjZ) (a,\cureff) \Rightarrow \vis{a}{\cureff}
\end{smathpar}

This ability to represent store semantics and application-level consistency in
the same language is vital to contract classification. Observe that out
contract language does not incorporate real (i.e., wall-clock) time. Hence, the
contract language cannot describe store semantics based on real time such as
recency or bounded-staleness guarantees offered by certain
stores~\cite{Pileus}.

\subsection{Contract Comparison}

Our goal is to classify the contracts, and map the corresponding operation to
the \emph{weakest} store level consistency. To this end, we need to be able to
compare the strengths of contracts. Let $\cv_{op}$ be a contract for a
particular operation $op$, and $\cv_{st}$ capture a particular store
consistency level. We would like to determine whether $op$ can be \emph{safely
discharged} at the store consistency level $\cv_{st}$ such that the resulting
execution does not violate $\cv_{op}$.

Since our contracts represent axiomatic definition of program executions, let
$\Mod{\cv}$ be the set of all executions under which $\cv$ is satisfied. If
every execution $\E \in \Mod{\cv_{st}}$ is also a member of $\Mod{\cv_{op}}$,
then $op$ can be safely discharged under the store consistency level
$\cv_{st}$. Formally, $\Mod{\cv_{st}} \subseteq \Mod{\cv_{op}}$. This is the
model-theoretic consequence relation, written as $\cv_{st} \models_m \cv_{op}$.

Observe that our contract language (stripped of its syntactic sugar) is a
carefully chosen subset of first-order logic that is known to be
decidable~\cite{epr}. Due to the completeness theorem~\cite{completeness},
$\cv_{st} \models_m \cv_{op}$ if and only if $\cv_{st} \Rightarrow \cv_{op}$.
Due to the decidability of our contract language this implication check is
automatically discharged with the help of a theorem prover. In this case, we
say that $\cv_{op}$ is \emph{weaker than} $\cv_{st}$ (written $\cv_{op} \le
\cv_{st}$). Formally,

\begin{definition}
An contract $\cv_{op}$ is said to be weaker than $\cv_{st}$ (written $\cv_{op}
\le \cv_{st}$ ) if and only if $\Delta \vdash \cv_{st} \Rightarrow \cv_{op}$.
\begin{center}
\end{center}
\end{definition}
\vspace{-1em}

\noindent where $\Delta$ captures assumptions about the nature of primitive
relations, such as $\Rvis$ and $\Rso$ are irreflexive, and the happens-before
relation $\hbZ$ is acyclic, preventing thin-air reads. \KC{Ideally, we need to
list all the axioms somewhere.}

@@

While the store semantics offered by commercial and research data stores vary
widely, for our exposition, we identify three particular consistency levels --
eventual, causal and strong, commonly offered by many distributed store with
tunable consistency, with increasing overhead in terms of latency and
availability. Indeed, the techniques presented here can be extended to other
consistency stratifications in a straight forward manner. We assign each
application-level contract into one of these following classes:

\begin{itemize}
\setlength{\itemsep}{2pt}

\item \textbf{Eventually consistency}: Eventually consistent operations can be
satisfied as long as the client can reach at least one replica. For example,
\cf{deposit} is a highly-available operation. While eventually consistent data
store typically offer \emph{basic} eventual consistency with all possible
anomalies, we assume that our store provides stronger semantics that remains
highly-available~\cite{BailisHAT,COPS}; the store always exposes causal cut of
the updates. This semantics can be formally captured as:

\vspace{-1em}
\begin{smathpar}
\ecc = \forall a,b. \hbo{a}{b} \wedge \vis{b}{\cureff} \Rightarrow \vis{a}{\cureff}
\end{smathpar}

\noindent where $\small \hboZ = ((\soZ \cap \sameobjZ) \cup \visZ)^+$.

\item \textbf{Causal consistency}: Operations with causally consistent
contracts are required to see a causally consistent snapshot of the object
state, including the actions performed on the same session.  The latter
requirement entails that if two operations $o_1$ and $o_2$ from the same
session are applied to two different replicas $r_1$ and $r_2$, the second
operation cannot be discharged until the effect of $o_1$ is merged with $r_2$.
We assume that causality is only tracked through operations on the same object;
two operations in the same session but on different objects are considered
causally unrelated under this definition. Stores typically avoid tracking
causaility across objects to mitigate the overhead due to unintended causality.
The corresponding store semantics is captured by the contract $\ccc$ presented
in Section~\ref{sec:store_sem}.

\item \textbf{Strong Consistency}: Strongly consistent operations may be block
indefinitely under network partitions. An example is the total-order contract
on \cf{withdraw} operation. The corresponding store semantics is captured by
the contract $\scc$ presented in Section~\ref{sec:store_sem}.

\end{itemize}

Observe that the contracts $\scc$, $\ccc$ and $\ecc$ are themselves totally
ordered with respect to the $\le$ relation: $\ecc \le \ccc \le \scc$. This
concurs with the intuition that any contract satisfiable under $\ecc$ or $\ccc$
is satisfiable under $\scc$ , and any contract that is satisfiable under $\ecc$
is satisfiable under $\ccc$. Nonetheless, we determine the class of a contract
based on the \emph{weakest} guarantee (among $\ecc$, $\ccc$, and $\scc$)
required to satisfy the contract. The classification scheme is presented
formally as rules in Figure~\ref{sem:classify}.

% Contract classification rules
% ------------------------------
\input{contract-classification}

The rules to classify contracts into availability classes are straightforward.
The remaining rule judges a contract as {\sf WellFormed} if and only if it can
be satisfied under strong consistency. The rest are reported as errors.
